---
title: "Regression tests"
---

```{r}
#| include: false
library(reticulate)
use_condaenv("stars-testing-intro", required = TRUE)
```

{{< include /assets/language-selector.html >}}

<br>

## What is a regression test?

A **regression** test involves running your workflow on historical data and confirming that results are consistent over time.

It's not focused on whether results are theoretically correct. It's about **consistency and reproducibility** - confirming that code changes, environment updates, or data pipeline tweaks have not silently changed results.


## Example: waiting times case study

We will run regression tests using the [dataset we introduced for our waiting times case study](case_study.qmd).

::: {.python-content}

In our test script, we need to import:

```{python}
#| eval: false
#| file: code/test_regression__imports.py
```

:::

## Regression test

::: {.python-content}

```{python}
#| eval: false
#| file: code/test_regression__test_reproduction.py
```

:::

::: {.r-content}

```{r}
#| eval: false
#| file: code/test_regression__re_running_on_historical.R
```

:::

## Running our example test

:::: {.callout-note title="Test output"}

::: {.python-content}

```{python}
#| echo: false
import pytest
pytest.main(["../examples/python_package/tests/test_regression.py"])
```

:::

::: {.r-content}

```{r}
#| echo: false
#| output: false
devtools::load_all("../examples/r_package")
```

```{r}
#| echo: false
testthat::test_file(
  "../examples/r_package/tests/testthat/test_regression.R"
)
```

:::

::::

## When should you update your regression tests?

There are two situations where you should deliberately update the expected results in a regression test.

**Bug fixes** - when you fix a real bug in the pipeline, you correct the code, then update the stored expected results so the new baseline reflects the bug fix and nothing else.

**Deliberate method changes** - when you intentionally change or improve the method itself, you also refresh the expected outputs so that any future failures signal behaviour that has changed beyond what you intended.

It's important to **ensure that any change in results comes from the bug fix or deliberate method update**, and not from an unintended change somewhere else.